---
title: "Reading & processing .TIF drone files"
output: html_document
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

The goal of this vignette is to show the process behind the function `read_n_process_tif`. This function allows you to read a `.TIF` file of a drone flight (obtained by processing drone imaging using the **Pix4D** software) and process it into a `tibble`, a `data.frame` data structure that makes this data easy to manipulate with `R`. To successfully use this function you will need to install the following packages in your system:

```{r, eval = FALSE}
library(tidyverse)
library(sf)
library(tiff)
library(raster)
library(proj4)
```

Luckily, these packages are included as imports for the `throne` package so once you install and load it, they will be automatically installed and loaded with it.

### Step 1: **Defining file path and processing information**

First thing is to define the location of the `.TIF` file you want to process. This kind of files are usually pretty big and exceed Git Hub's maximum file size limit so we recommend keeping them in a separate `data` folder. In my case, this is the file path:

```{r}
filepath <- "C:/Users/ggarc/OneDrive/projects/throne_data/test_flight.tif"
```

In addition, we you should also define the other `3` inputs of the function. First is the **Universal Transverse Mercator** (`UTM`) zone in which your drone flight took place. The `UTM` coordinate system divides the world into 60 north-south zones, each 6 degrees of longitude wide. To calculate which `UTM` zone your flight took place, you can use the rough longitude of where your flight took place and apply the formula:

$$
UTM_{zone} = \lfloor \frac{Longitude + 180}{6} \rfloor + 1
$$

In my case, the average longitude of the flight used as an example here was -119.45 so the `UTM` zone would then be:

```{r}
utm_zone = floor((-119.45 + 180) / 6) + 1
utm_zone
```

Next we need to define the hemisphere in which the flight took place and that will be either `north` or `south`. In our case, the example flight took place in the northern hemisphere:

```{r}
hemisphere <- "north"
```

Lastly, you will need to define the number of decimal digits that of your coordinate. Using a decimal coordinate system (i.e., -119.45 instead of other coordinate systems like the DMS as 119° 27' 0" W which we highly discourage) each decimal degree will convey a certain distance. These distances differ between longitude and latitude degrees and will become smaller at higher latitudes. As part of the documentation of this package we have included a data set (`spatial_resolution_lat_long.csv`) to visualize this pattern. Below is a plot generated with that data to express the area of a rectangle in which each of the sides (representing longitude and latitude) has a specific number of decimal digits (i.e., 1, 0.1, 0.001 etc.) at different latitudes.

```{r, echo = FALSE, message = FALSE, fig.width=4, fig.height=4, fig.align='center'}
# load spatial data 
library(readr)
library(tidyverse)
spr_data <- read_csv("C:/Users/ggarc/OneDrive/projects/throne/data/spatial_resolution_lat_long.csv")

# process data for plot
spr_data_plot <- spr_data[,1:5] %>% 
  pivot_wider(names_from = long_lat, values_from = spatial_resolution) %>%
  mutate(area = longitude * latitude) 

# plot all decimal degrees
spr_data_plot %>% 
  ggplot(aes(x = decimal_places, y = area,col = as.factor(deg_latitude))) +
  geom_line(lwd = 2, alpha = 0.75) +
  scale_x_continuous(breaks = c(0:8)) +
  xlab("Decimal places") +
  ylab("Spatial resolution (m^2)") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.7)) +
  guides(color = guide_legend(title = "Degrees Latitude"))
```

As seen below, we recommend a spatial resolution of **5 decimal places** since, depending on latitude, will convey an area smaller than 1 m^2 which is a relevant scale for the vast majority of terrestrial ectotherms:

```{r, echo = FALSE, message = FALSE, fig.width=4, fig.height=4, fig.align='center'}
# plot > 4 decimal degrees
spr_data_plot %>% 
  filter(decimal_places > 3) %>%
  ggplot(aes(x = decimal_places, y = area,col = as.factor(deg_latitude))) +
  geom_line(lwd = 2, alpha = 0.75) +
  scale_x_continuous(breaks = c(0:8)) +
  xlab("Decimal places") +
  ylab("Spatial resolution (m^2)") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.7)) +
  guides(color = guide_legend(title = "Degrees Latitude"))
```

```{r}
decimal_places <- 5
```

### Step 2: **Reading `.TIF` file and transform into a `tibble` object**

Using the file path defined earlier, we can use the `raster` function from the package of the same name in order to read the file into your `R` environment:

```{r}
raster <- raster(filepath)
```

The file will load in as a `Formal class RasterLayer` which can actually be plotted using base `R`:

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
plot(raster)
```

In order to have a more easy to work with this data, we can use the `rasterToPoints` function from the `raster` package to transform our raster object into a data frame.

```{r}
# transform into points
points <- rasterToPoints(raster)

# transform into data frame
df <- as.data.frame(points)

as_tibble(df)
```

The resulting data frame will have `3` columns. First there is `x` and `y`, which are equivalent to `longitude` and `latitude` (although in different units, more on that down below) and a column called `layer` which is equivalent to the temperature the drone captured in degrees Celsius. The number of rows of this data frame will be heavily dependent on the resolution of your drone's camera. The example file was obtained using **THIS DRONE** with **THIS CAMERA** which is a bit of an overkill. A 10th of observations would have worked just fine.

Before moving on to the next steps, it is important to note that the majority of observations on the `layer` column are `0`. This is not because the temperature recorded by the drone was exactly `0` but because the drone did not record data. If these `0` observations are removed that reduces the size of the final data frame substantially. 

```{r}
# removing layer = 0 observations
df <- df %>% filter(layer != 0) 

as_tibble(df)
```





