---
title: "Reading & processing flight data"
output: html_document
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

The goal of this vignette is to show the process behind the function `read_n_process_flight_data`. This function allows you to read raw data from a drone flight in a `.TIF` file format obtained by processing the images gathered during the flight itself. The function will read and processes the `.TIF` file into a `tibble`, a `data.frame`-like data structure that makes this data easy to manipulate with `R`. To successfully use this function you will need to install the following packages in your system:

```{r, eval = FALSE}
library(tidyverse)
library(sf)
library(tiff)
library(raster)
library(proj4)
```

Luckily, these packages are included as dependencies for the `throne` package so once you install and load it, they will be automatically installed and loaded with it.

### Step 1: **Defining file path and processing information**

First thing is to define the location of the `.TIF` file you want to process. This kind of files are usually pretty big and exceed GitHub's maximum file size limit so keep that in mind when managing your data. In our case, we stored flight information in a separate folder locally. This was our file path:

```{r}
filepath <- "C:/Users/ggarc/OneDrive/projects/throne_data/test_flight.tif"
```

Before continuing, we you should also define the other `3` inputs of the `read_n_process_flight_data` function. First is the **Universal Transverse Mercator** (`UTM`) zone in which your drone flight took place. The `UTM` coordinate system divides the world into `60` north-south zones, each `6` degrees of longitude wide. To calculate which `UTM` zone your flight took place, you can use the rough longitude of where your flight took place and apply the formula:

$$
UTM_{zone} = \lfloor \frac{Longitude + 180}{6} \rfloor + 1
$$
In my case, the average longitude of the flight used as an example here was `-119.45` so the `UTM` zone would then be:

```{r}
utm_zone = floor((-119.45 + 180) / 6) + 1
utm_zone
```

The `throne` package also includes the `get_utm_zone` function which will perform this calculation for you. To use it, you can input a rough `longitude` of where your flight took place but be aware that this value needs to be between `-180` and `180` because, unfortunately, the Earth is **NOT** flat.

```{r}
get_utm_zone(15.4)
get_utm_zone(195)
```

Next we need to define the hemisphere in which the flight took place; either `north` or `south`. In our case, the example flight took place in the northern hemisphere:

```{r}
hemisphere <- "north"
```

Lastly, you will need to define the number of decimal digits that of your coordinate. Using a decimal coordinate system (i.e., `-119.45` instead of other coordinate systems like the DMS as `119° 27' 0" W` which we highly discourage) each decimal degree will convey a certain distance. These distances differ between longitude and latitude degrees and will become smaller at higher latitudes. As part of the documentation of this package we have included a data set (`spatial_resolution_lat_long.csv`) to visualize this pattern. Below is a plot generated with that data to express the area of a rectangle in which each of the sides (representing longitude and latitude) has a specific number of decimal digits (i.e., `1`, `0.1`, `0.001` etc.) at different latitudes.

```{r, echo = FALSE, message = FALSE, fig.width=4, fig.height=4.5, fig.align='center'}
# load spatial data 
library(readr)
spr_data <- read_csv("C:/Users/ggarc/OneDrive/projects/throne/data/spatial_resolution_lat_long.csv")

# process data for plot
spr_data_plot <- spr_data[,1:5] %>% 
  pivot_wider(names_from = long_lat, values_from = spatial_resolution) %>%
  mutate(area = longitude * latitude) 

# plot all decimal degrees
spr_data_plot %>% 
  ggplot(aes(x = decimal_places, y = area,col = as.factor(deg_latitude))) +
  geom_line(linewidth = 2, alpha = 0.75) +
  scale_x_continuous(breaks = c(0:8)) +
  xlab("Decimal places") +
  ylab("Spatial resolution (m^2)") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.7)) +
  guides(color = guide_legend(title = "Degrees Latitude"))
```

We recommend a spatial resolution of **`5` decimal places** since, for most latitudes, it will convey an area $< 1 m^2$ which is a relevant scale for the vast majority of terrestrial animals.

```{r, echo = FALSE, message = FALSE, fig.width=4, fig.height=4, fig.align='center'}
# plot > 4 decimal degrees
spr_data_plot %>% 
  filter(decimal_places > 4) %>%
  ggplot(aes(x = decimal_places, y = area,col = as.factor(deg_latitude))) +
  geom_line(lwd = 2, alpha = 0.75) +
  scale_x_continuous(breaks = c(0:8)) +
  xlab("Decimal places") +
  ylab("Spatial resolution (m^2)") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.7)) +
  guides(color = guide_legend(title = "Degrees Latitude"))
```

```{r}
decimal_places <- 5
```

### Step 2: **Reading the flight's file (`.TIF`) and transform into a `tibble` object**

Using the file path defined earlier, we can use the `raster` function from the package of the same name in order to read the file into your `R` environment:

```{r}
raster <- raster(filepath)
```

The file will load in as a `Formal class RasterLayer` which can actually be plotted using base `R`:

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
plot(raster)
```

Then, can use the `rasterToPoints` function from the `raster` package to transform our raster object into a `data.frame`:

```{r}
# transform into points
points <- rasterToPoints(raster)

# transform into data frame
df <- as.data.frame(points)

as_tibble(df)
```

The resulting `data.frame` will have `3` columns. First there is `x` and `y`, which are equivalent to `longitude` and `latitude` (although in different units, more on that on **Step 3**) and a column called `layer` which is equivalent to the temperature the drone captured in degrees Celsius. The number of rows of this data frame will be heavily dependent on the resolution of your drone's camera. The example flight was obtained using **DJI Matrice 200 V2** with a **FLIR Zenmuse XT2** thermal imaging camera which is a bit of an overkill. A 10th of observations would have worked just fine.

It is important to note that the majority of observations on the `layer` column are `0`. This is not because the temperature recorded by the drone was exactly `0` but because the drone did not record data. If these `0` observations are removed that reduces the size of the final `data.frame` (presented as a `tibble` below) substantially: 

```{r}
# removing layer = 0 observations
df <- df %>% filter(layer != 0) 

as_tibble(df)
```

### Step 3: **Transforming coordinates and generating final output**

As pointed out earlier, the coordinate columns of the final `tibble` obtained in the previous step are in a weird format. Currently, `x` and `y` are defined as the distance in meters to the western and northern edges of the `UTM` zone where the flight took place. In other to correlate this information with data from other sources, you will need to transform this data into a **decimal coordinate system**. To do so, we first need to define a **projection string**, which will provide the information to the function `project` from the `proj4` package to perform this transformation. The `read_n_process_flight_data` function will do this for you, but here is what is happening behind the scenes:

```{r}
# generate the string including utm_zone and hemisphere
string <-  paste("+proj=utm +zone=", utm_zone," +", hemisphere," +ellps=WGS84 +datum=WGS84 +units=m +no_defs", sep = "")
string
```
We can then `project` our data into decimal coordinate units. This is the computationally longest step of the entire process.

```{r}
# project to decimal coordinates
coords <- as.data.frame(project(data.frame(x = df$x, y = df$y), string, inverse = TRUE))
head(coords)
```

Last but not least we need to define the spatial resolution at which we are conducting our measurements which will be dependent on the amount of `decimal_places` we defined earlier. Once more the `read_n_process_flight_data` function will do this for you, but here is what is happening behind the scenes:

```{r}
# add temperature column to coordinates 
coords$temp <- df$layer

# rename x & y columns 
data <- coords %>% rename(long = x, lat = y)

# define decimal degrees on latitude and longitude measurements
data$long <- as.numeric(format(round(data$long, decimal_places), nsmall = decimal_places))
data$lat <- as.numeric(format(round(data$lat, decimal_places), nsmall = decimal_places))

# get mean temperature for same coordinates 
data <- data %>% group_by(lat, long) %>% summarise(temp = mean(temp, na.rm = T)) %>% ungroup()
data
```

The last summarizing step reduces the size of the data frame to the point where this data can already be plotted using `ggplot` tools and be part of any posterior analysis. For instance, we can visualize it as a map using `geom_tile` or `geom_raster`: 

```{r, fig.width = 6, fig.height = 6, fig.align='center'}
data %>% 
  ggplot(aes(x = long, y = lat, fill = temp)) +
  geom_raster() +
  scale_fill_gradient(low = "lightgoldenrod1", high = "red4") +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(legend.position = c(0.95, 0.85))
  
```
Note that the image plotted here is exactly the same as the one showed earlier when plotting the `raster` but altering the color scheme. Additionally, we can already get some descriptive metrics of the thermal landscape which we can visualize as a distribution: 

```{r}
data %>% 
  ungroup() %>%
  ggplot(aes(x = temp)) +
  geom_density(fill = "lightgoldenrod", linewidth = 1.25) +
  geom_vline(aes(xintercept = mean(temp, na.rm = T)), linewidth = 1.25) +
  xlab("Temperature (as captured by drone,°C)") + ylab("Density") +
  theme_minimal() 
```



